---
title: "bayesian lab1"
output: html_document
date: "2026-04-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. a)

```{r}
s <- 43
f <- 35
n <- 10000
alpha <- 7 + s
beta <- 7 + f

set.seed(123)
X <- rbeta(n, shape1 = alpha, shape2 = beta)

true_mu <- alpha/(alpha + beta)
true_sd <- sqrt(alpha*beta/((alpha + beta)^2 * (alpha + beta + 1)))
mu <- c()
sd <- c()
for (i in 1:n) {
  mu <- c(mu, mean(X[1:i]))
  sd <- c(sd, sd(X[1:i]))
  
}

curve(dbeta(x, shape1 = alpha, shape2 = beta))

plot(mu, t = "l")

abline(h = true_mu, col = "red")

plot(sd, t = "l")

abline(h = true_sd, col = "red")

```
### b)
```{r}
1-pbeta(0.5, shape1 = alpha, shape2 = beta)
mean(X>0.5)

```
### c)
```{r}
hist(X/(1-X))

```


## 2.
```{r}
library(LaplacesDemon)
y <- c(22, 33, 31, 49, 65, 78, 17, 24)
mu <- 3.65
n <- length(y)

tao <- sum((log(y)-mu)^2) / n

X <- rinvchisq(10000, n, tao)
curve(dinvchisq(x, n, tao), from = 0.001, to = 3)

```
### b)

```{r}
library(bayestestR)


Gini <- 2 * pnorm(sqrt(X)/sqrt(2), 0, 1) - 1
print("Credible interval:")
print(quantile(Gini, c(0.025, 0.975)) |> round(3))

print("HPDI:")
HPDI <- hdi(Gini)
HPDI <- c(HPDI$CI_low, HPDI$CI_high)
print(HPDI |> round(3))

hist(Gini)
abline(v = quantile(Gini, c(0.025, 0.975)), col = "blue")
legend("topright", 
       legend = c("Credible Interval", "HPDI"), 
       fill = c("blue", "red"))

abline(v = HPDI, col = "red")



```

## 3.
### a)
As the data points were independent observations from the Poisson distribution, the likelihood was 

$$
p(y \mid \lambda)
= \prod_{i=1}^n \frac{e^{-\lambda}\lambda^{y_i}}{y_i!}
$$
which was then simplified and constants not depending on $\lambda$ were dropped:

$$
p(y \mid \lambda)
= \prod_{i=1}^n \frac{e^{-\lambda}\lambda^{y_i}}{y_i!} = \frac{e^{-\lambda n}\lambda^{\sum y_i}}{\prod y_i!} \propto e^{-\lambda n}\lambda^{\sum y_i}
$$
Similarly, constants not depending on $\lambda$ in the prior were dropped:

$$
p(\lambda \mid \sigma)
=
\frac{\sqrt{2}}{\sigma\sqrt{\pi}}
\exp\!\left(-\frac{\lambda^2}{2\sigma^2}\right)\propto e^{-\frac {\lambda^2}{2\sigma^2}},
\quad \lambda \ge 0
$$


The posterior then, under Bayesian inference, becomes

$$
p(\lambda \mid y) \propto p(y \mid \lambda) \cdot p(\lambda \mid \sigma) 
$$
Using the expressions derived above, we obtain

$$
p(\lambda \mid y) \propto e^{-\lambda n}\lambda^{\sum y_i} \cdot e^{-\frac {\lambda^2}{2\sigma^2}} = e^{(-\lambda n - \frac {\lambda^2}{2\sigma^2} )}\lambda^{\sum y_i}
$$

Substituting $n=7$, $\sigma=5$ and $\sum y_i = 24$, we get

$$
p(\lambda \mid y) \propto e^{(-7\lambda  - \frac {\lambda^2}{50})}\lambda^{24}
$$
Since the normalizing constant does not admit a closed-form expression, the posterior distribution was normalized numerically to integrate to one.

$$
p(\lambda \mid y)
=
\frac{
\lambda^{24}\exp\!\left(-7\lambda - \frac{\lambda^2}{50}\right)
}{
\int_0^\infty
\lambda^{24}\exp\!\left(-7\lambda - \frac{\lambda^2}{50}\right)\, d\lambda
},
\quad \lambda \ge 0
$$
The below plot shows the normalized posterior distribution of the average goals rate parameter.

```{r echo=FALSE}
lambda_grid <- seq(0, 8, length.out = 2000)

unnorm_post <- lambda_grid^24 *
  exp(-7 * lambda_grid - lambda_grid^2 / 50)

delta <- lambda_grid[2] - lambda_grid[1]
Z <- sum(unnorm_post) * delta

post <- unnorm_post / Z

plot(lambda_grid, post, type = "l", lwd = 2, col ="red",
     xlab = expression(lambda~"(average goals per match)"),
     ylab = "Posterior density",
     main ="Posterior Distribution of the Average Goals Rate")
```


### b)
From the information in *a*, the posterior mode of $\lambda$ was obtained.

```{r echo=FALSE}
lambda_mode <- lambda_grid[which.max(post)]
lambda_mode
```

## APPENDIX
```{r eval=FALSE}
#3
#a
lambda_grid <- seq(0, 8, length.out = 2000)

unnorm_post <- lambda_grid^24 *
  exp(-7 * lambda_grid - lambda_grid^2 / 50)

delta <- lambda_grid[2] - lambda_grid[1]
Z <- sum(unnorm_post) * delta

post <- unnorm_post / Z

plot(lambda_grid, post, type = "l", lwd = 2, col ="red",
     xlab = expression(lambda~"(average goals per match)"),
     ylab = "Posterior density",
     main ="Posterior Distribution of the Average Goals Rate")

#b
lambda_mode <- lambda_grid[which.max(post)]
lambda_mode
```